---
title: "Evolution of demographic parameters and population stability"
output:
  html_notebook: default
  pdf_document: default
---
Computational biology project by Bianca BOI and Anne STAFF

## Some plots:
```{r include=FALSE}
gen <- read.table("genz.txt")
gen$r <- gen$N / gen$K_means
x <- nrow(gen)
```
\newpage
```{r}
plot(1:x, gen$N, type="b", pch=18, col="red", xlab="Number of generations", ylab="population size", main="Population size over generations")
```
```{r}
plot(24890:24898, gen$N[24890:24898], type="b", pch=18, col="red", xlab="Number of generations", ylab="population size", main="Population size over generations")
```
```{r}
gen[24890:24898,]
```


```{r}
plot(1:24880, gen$N[1:24880], type="b", pch=18, col="red", xlab="Number of generations", ylab="population size", main="Population size over generations")
```

\newpage
```{r}
# plot the relative density of population with mutation
plot(1:x, gen$r, type="b", pch=20, col="blue", xlab="Time in generations", ylab="relative density N/mean(K)", main="Evolution of the relative density over generations")
```
```{r}
# plot the relative density of population with mutation
plot(1:24880, gen$r[1:24880], type="b", pch=20, col="blue", xlab="Time in generations", ylab="relative density N/mean(K)", main="Evolution of the relative density over generations")
```
Comparing with the paper, we noticed that the relative density should eventually become equal to 1 for large times. Unfortunately, over the small time span we covered, this is not the case for us.

\newpage
This function plots the mean values of the traits we monitored along with their corresponding standard deviation:
```{r}
plot_mean_with_sd <- function(means, 
                              sd, 
                              interval=NULL, 
                              xlab="Time in generations", 
                              ylab=NULL, 
                              main=NULL) 
{
  if (is.null(interval)) {
    interval <- 1:length(means)
  }
  plot(interval, 
       means[interval], 
       type = 'l', 
       col="blue", 
       ylim=c(min(means-sd), max(means+sd)), 
       xlab=xlab, 
       ylab=ylab, 
       main=main)
  
  polygon(c(rev(interval), interval), 
          c(rev(means[interval]-sd[interval]),
            means[interval]+sd[interval]), 
          col = 'grey80', border = NA)
  
  lines(interval, means[interval], col="blue")
  lines(interval, means[interval]-sd[interval], lty = 'dashed', col = 'red')
  lines(interval, means[interval]+sd[interval], lty = 'dashed', col = 'red')
}
```

\newpage
```{r}
plot_mean_with_sd(gen$K_means, 
                  gen$K_sd, 
                  ylab="mean(K) \u00B1 sd(K)", 
                  main="Mean equilibrium density per generation and their standard deviation")
```
```{r}
plot_mean_with_sd(gen$K_means, 
                  gen$K_sd, 
                  interval = 24880:24950,
                  ylab="mean(K) \u00B1 sd(K)", 
                  main="Mean equilibrium density per generation and their standard deviation")
```

\newpage
```{r}
plot_mean_with_sd(gen$c_means, 
                  gen$c_sd, 
                  ylab="mean(c) \u00B1 sd(c)", 
                  main="Mean complexity per generation and their standard deviation")
```


```{r}
plot_mean_with_sd(gen$c_means, 
                  gen$c_sd, 
                  interval=24880:24900, 
                  ylab="mean(c) \u00B1 sd(c)", 
                  main="Mean complexity per generation and their standard deviation")
```

```{r}
plot_mean_with_sd(gen$c_means, 
                  gen$c_sd, 
                  interval=24894:24898, 
                  ylab="mean(c) \u00B1 sd(c)", 
                  main="Mean complexity per generation and their standard deviation")
```
\newpage
```{r}
plot_mean_with_sd(gen$c_means, 
                  gen$c_sd, 
                  ylab="mean(lambda) \u00B1 sd(lambda)", 
                  main="Mean growth rate per generation and their standard deviation")
```

```{r}
plot_mean_with_sd(gen$c_means, 
                  gen$c_sd, 
                  interval = 24880:24900,
                  ylab="mean(lambda) \u00B1 sd(lambda)", 
                  main="Mean growth rate per generation and their standard deviation")
```

```{r}
plot_mean_with_sd(gen$c_means, 
                  gen$c_sd, 
                  interval = 24894:24898,
                  ylab="mean(lambda) \u00B1 sd(lambda)", 
                  main="Mean growth rate per generation and their standard deviation")
```

\newpage
```{r}
plot(1:x, gen$fitness_means, type = 'l', col="blue", xlab="Time in generations", ylab="mean(fitness)", main="Mean fitness per generation")
```

```{r}
plot(1:24880, gen$fitness_means[1:24880], type = 'l', col="blue", xlab="Time in generations", ylab="mean(fitness)", main="Mean fitness per generation")
```
